---
title: "sncp: an R package for fitting a Shot Noise Cox Process to lung CT scans"
author: "Brian Vestal"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
# Introduction
  This vignette goes through an example of using the functions available in the sncp R package to get estimated parameters for a Shot Noise Cox Process (SNCP) fit to lung CT scans that show radiologically based evidence of emphysema as described in "Using a Spatial Point Process Framework to Characterize Lung Computed Tomography Scans" by Vestal et al.  We begin by loading in the sncp R package, along with several others that will be utilized for model summaries and visualizations.
  
```{r, warning = FALSE, message = FALSE}
  library(sncp)
  library(ggplot2)
  library(dplyr)
  library(ellipse)
```
  
Next, we visualize the simulated point pattern:
```{r}
ggplot(boundry_total) +
  aes(x = x, y = y) +
  geom_tile() +
  geom_tile(data = sim_data$pointPattern, color = 'red') +
  geom_point(data = sim_data$centers, color = 'blue') +
  ggtitle("Simulated Point Pattern")
```

Here we can see that there are 10 true clusters (centers are the blue dots) with the observed pixels (e.g. emphysema) as the red tiles.

# Fitting the SNCP 
There are two versions of the SNCP that we can fit where the only difference is how new points are proposed in the birth-death process.  In sncp_bdmcmc we use a uniform proposal distribution, where as in sncp_bdmcmc_smooth we use the data driven proposal surface described in the above referenced paper.  The input arguments are the same for both with the exception of the sncp_bdmcmc_smooth version needing the smoothing bandwidth for the proposal surface as well.  In this example, we will use the sncp_bdmcmc_smooth version (should take a few seconds to run):

```{r}
resultsBDMCMC <- sncp_bdmcmc_smooth(obs_points = as.matrix(sim_data$pointPattern),
                                    mean_mu_alpha = 8,
                                    sd_log_alpha = sqrt(3),
                                    sd_prop_alpha = .5,
                                    beta = .001,
                                    n_it = 10000,
                                    window_hw = 5,
                                    df_iw_prior = 5,
                                    df_iw_prop = 15,
                                    sigma_prior = diag(x = 40, nrow = 2),
                                    lung_data = as.matrix(lung_data),
                                    var_mu_alpha = 5,
                                    pen_dist = 20,
                                    pen_val = 1e-9,
                                    n_cent_init = 2,
                                    prior_n_cent = 3,
                                    max_bd_events = 20,
                                    max_bd_vt = 5,
                                    sigma_smooth = 100)
```

# Summarizing the Results
Now that we have the posterior estimates for each of the model parameters, we can explore how well they match the true values that went into generating the simulated data set.  

##  Number of clusters
We begin by examining the number of clusters that are estimated to be there, along with their location:

```{r}
###   Heat map of center locations

##  Collapsing the estimated cluster centers into a single data frame
centers_list <- lapply(resultsBDMCMC$centers_sample, function(x){
  ret <- as.data.frame(x)
  names(ret) <- c("x", "y")
  return(ret)
})
centers_sample_collapsed <-  bind_rows(centers_list)

##  Getting rounded x and y coordinates for building the heatmap at the pixel level
centers_sample_collapsed$y_round <- round(centers_sample_collapsed$y)
centers_sample_collapsed$x_round <- round(centers_sample_collapsed$x)

##  Getting counts for each pixel of how many iterations it was estimated to be a cluster center
centers_heat_map_df <- centers_sample_collapsed %>%
  group_by(x_round, y_round) %>%
  summarise(num = n()) %>% 
  filter(num > 5) #only plotting pixels that show up in at least 5 iterations 

names(centers_heat_map_df)[1:2] <- c('x', 'y')
ggplot(centers_heat_map_df) +
  aes(x = x, y = y) +
  geom_tile(aes(fill = num), height = 1, width = 1) +
  scale_fill_gradient(low = "blue", high = "red") +
  geom_tile(data = boundry_total, fill = "black", aes(x = x, y = y)) +
  xlab("x") +
  ylab("y") +
  ggtitle("Posterior distribution of \nestimated cluster centers")
```

Next we look at how many clusters there were estimated to be by using the posterior median, and we examine the MCMC's mixing and convergence by creating a trace plot of this value:

```{r}
df_n_cent <- data.frame(num_centers = resultsBDMCMC$n_centers_sample, 
                        it =1:length(resultsBDMCMC$mu_alpha_sample))
ggplot(df_n_cent) +
  aes(x = it, y = num_centers) +
  geom_path() +
  ggtitle("Trace plot for number \nof estimated centers")
summary(resultsBDMCMC$n_centers_sample)

### Estimation error:
median(resultsBDMCMC$n_centers_sample) - sim_data$nClust
```

Now we will create a plot that shows the estimated clustering from the last iteration of the BD-MCMC


```{r}
ellipseLevel <- .75

centers_sub <- data.frame(resultsBDMCMC$centers_sample[[10000]])
names(centers_sub) <- c('x', 'y')
sigmas_sub <- resultsBDMCMC$sigmas_sample[[10000]]
ellipse_tmp <- do.call(rbind, lapply(1:nrow(centers_sub), function(j){
  dat_tmp <- ellipse::ellipse(x = sigmas_sub[, , j], level = ellipseLevel, centre = as.numeric(centers_sub[j, ]))
  ret <- cbind(data.frame(dat_tmp), cluster = j)
  return(ret)
}))
ggplot(boundry_total) +
  aes(x = x, y = y) +
  geom_tile() +
  geom_tile(data = sim_data$pointPattern, fill = 'red') +
  geom_point(data = centers_sub, color = 'black', size = 1) +
  geom_path(data = ellipse_tmp, aes(group = cluster)) +
  ggtitle("Estimated Clustering")

```

## Average Cluster Size

Next we look at the posterior distribution for the average size of clusters (on the log scale).  

```{r}
###   Trace plot for mu_alpha
tp_mu_alpha_df <- data.frame(mu_alpha = resultsBDMCMC$mu_alpha_sample,
                             it = 1:length(resultsBDMCMC$mu_alpha_sample))

ggplot(tp_mu_alpha_df) +
  aes(x = it, y = mu_alpha) +
  geom_path() +
  ggtitle("Trace plot for mu_alpha")
```

From the trace plot we see that the sampler seems to have converged to the target distribution very quickly.

```{r}
###   Summary of mu_alpha posterior sample
summary(resultsBDMCMC$mu_alpha_sample)

###   Estimation error for mu_alpha using posterior median
median(resultsBDMCMC$mu_alpha_sample) - sim_data$muAlpha
```

## Measure of Random Scatter

Finally we look at the posterior distribution for the amount of random scatter, or diseased pixels that do not show evidence of clustering (again on the log scale).  

```{r}
###   Trace plot for eta
tp_eta_df <- data.frame(eta = resultsBDMCMC$beta_sample,
                             it = 1:length(resultsBDMCMC$beta_sample))

ggplot(tp_eta_df) +
  aes(x = it, y = eta) +
  geom_path() +
  ggtitle("Trace plot for eta")

###   Summary of log(eta) posterior sample
summary(resultsBDMCMC$beta_sample)

###   Estimation error for eta using posterior median
median(resultsBDMCMC$beta_sample) - sim_data$beta
```

